plyranges: a grammar for transforming genomics data

### 2018-05-08 paper.Rmd
more detail for bam example

bibliography: references.bib
output: 
  rticles::plos_article:
csl: plos.csl
---
```{r setup, include = FALSE}
library(tibble)
library(knitr)
opts_chunk$set(message = FALSE, 
               warning = FALSE, 
               comment = "#>", 
               fig.width = 5, 
               fig.height = 3, 
               fig.align = "center")
# for peak figures
library(ggbio)

library(plyranges)

@@

@@

Another common operation in genomics data analysis is to compute data
summaries over genomic windows. In `plyranges` this can be achieved via
the `group_by_overlaps` operator. Continuing with the data from the 
Human Epigenome Roadmap Consortium data, we can count the number of reads
that fall into a fixed bins of size 10000bp over a BAM file of H3K27Me3 methylation
marks from the H1 cell line. We extract reads that have a mapping quality
score greater than 20:

```{r, eval = FALSE}
alignments <- read_bam(h1_bam) %>%
  filter(mapq > 20) 
```

Note that the BAM file is only read into memory once we perform an operation
on it. Because of interoperability over Bioconductor, we can generate
genomic bins using the `tile` function from `GenomicRanges`:

```{r, eval = FALSE}
bins <- tile(h1_bam, width = 10000)
```

@@

```{r, eval = FALSE}
h1_bam_read_summary <- h1_bam %>%
  group_by_overlaps(bins) %>%
  summarise(n_reads = n())
```

>>>>

bibliography: references.bib
output: 
  rticles::plos_article:
csl: plos.csl
---
```{r setup, include = FALSE}
library(tibble)
library(knitr)
opts_chunk$set(message = FALSE, 
               warning = FALSE, 
               comment = "#>", 
               fig.width = 5, 
               fig.height = 3, 
               fig.align = "center")
# for peak figures
library(ggbio)
library(plyranges)

@@

h1_bam_sorted <- "./data/GSM433167_BI.H3K27me3.sorted.bam"

# sort and index bam
if (!file.exists(h1_bam_sorted)) {
  Rsamtools::sortBam(h1_bam, "./data/GSM433167_BI.H3K27me3.sorted")
  Rsamtools::indexBam(h1_bam_sorted)
}

@@

Another common operation in genomics data analysis is to compute data
summaries over genomic windows. In `plyranges` this can be achieved via
the `group_by_overlaps` operator. Continuing with the data from the 
Human Epigenome Roadmap Consortium data, we can count the number of reads
that fall into a fixed bins of size 10000bp over a sorted and indexed 
BAM file of H3K27Me3 methylation marks from the H1 cell line.

We can resuse the annotation information from our BigWig file example, as the
BAM file was aligned to the hg19 genome build. Alignments over
the mitochondrial genome are dropped using the `dropSeqlevels` function
from the Bioconductor package `GenomeInfoDb` @R-gdb.
```{r locations, eval = FALSE}
locations <- bw_file %>%
  get_genome_info() %>% 
  dropSeqlevels("chrM", pruning.mode = "coarse") %>%
  set_genome_info(genome = "hg19") 
```

To generate bins of width 10000bp we can use the `tile` function from 
`GenomicRanges`:

```{r, eval = FALSE}
bins <- tile(locations, width = 10000)
```

Next we only read in alignments if they have a map quality score
greater than 20.
```{r, eval = FALSE}
alignments <- read_bam(h1_bam_sorted) %>%
    filter(mapq > 20) 
```

@@

```{r, eval = FALSE}
alignments_summary <- bins %>%
  group_by_overlaps(alginments) %>%
  summarise(n_reads = n())
```
